Reference Tracking Strategies of Deaf Adult Signers in Turkish Sign Language

### 2023-01-31 revised_manuscript.Rmd
updated

---
editor_options:
  chunk_output_type: console
title: Reference Tracking Strategies of Native and Late Signers in Turkish Sign Language
abstract: |
  We investigated the reference tracking strategies among native and late adult deaf signers of Turkish Sign Language. Consistent with the theories of saliency and referential accessibility, regardless of their acquisiton groups, signers mainly used nominals and extension classifiers for introducing referents. To maintain a referent in the previous clause, signers chose null marking (e.g., constructed action, agreement and plain verbs). For topic shifts or re-introduced contexts, nominals and pronouns were chiefly favored although we observed very little pronominal use. As for age of acquisition effects, we only report limited over-redundancy for late signers who used null marking less compared to native signers especially for introduced and maintained contexts. Given our limited support for redundancy by late signers, we conclude that native signers might utilize the spatial affordances of the visual-spatial modality better than late signers only to a certain extent.
keywords: "Referential accessibility, signed narratives, pragmatic competence, delayed language acquisition, Turkish Sign Language"
journal: "Journal of Pragmatics"
date: "`r Sys.Date()`"
bibliography: ./bibliography/references.bib
times: true
numbersections: trueand w
classoption: "review"
csl: ./bibliography/elsevier-harvard.csl
output: rticles::elsevier_article
header-includes:
   - \usepackage{floatrow}
   - \usepackage{hyperref}
   - \usepackage{enumitem}  
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \floatplacement{table}{H}
   - \floatsetup[figure]{capposition=top}
   - \floatsetup[table]{capposition=top}
   - \usepackage{times}
   - \usepackage{smartdiagram}
   - \usepackage{ragged2e}
   - \usepackage{tikz}
   - \usepackage{xcolor}
   - \usepackage[utf8]{inputenc}
   - \usepackage[T1]{fontenc}
   - \usepackage{graphicx}
   - \usepackage{./sty/langsci-gb4e}
   - \usepackage{caption}
   - \usepackage{multicol}
   - \setlength{\columnsep}{-3cm}
   - \usepackage{setspace}
---

@@

Narratives are extensive discourses that construct an important part of human communication. In order to create a systematic and coherent organization of experiences, narrators must choose appropriate linguistic forms to refer to entities [@ariel1990; @givon1983; @gundel1993]: Certain forms such as nominal expressions are commonly used for the first mention of a referent in a discourse, and other forms such as pronouns and null marking may track previously mentioned entities which could either be maintained across multiple clauses or re-introduced back into the discourse. The selection of these referring expressions (henceforth REs) is contingent on cognitive abilities, such that individuals must rely on the activation of the working memory to keep in mind previously referred expressions and long strings of information [@morgan2005; @bamberg1997]. Pragmatic abilities too play a pivotal role in the tracking of referents. The form with the least amount of redundant information must be selected based on the accessibility levels of the referent (i.e., how active a referent is in the addressee's mind given a point in discourse) [@ahn2019; @ariel1990]. Because of these cognitive and pragmatic necessities, it is not surprising to observe that producing a coherent narrative is one of the latest milestones in language acquisition among children [@hickmann1996]. As can be expected, there is a plethora of research examining how narrative competence is achieved in both monolingual and bilingual acquisition contexts among children and adults. Although narrative competence and referent selection have been studied extensively for typically developing speakers [@aksu-koc2015; @hickmann1996; @williams1988], deaf individuals who acquired a sign language in atypical learning contexts (i.e., deaf signers with linguistic deprivation) have remained mostly understudied with an exception of handful of studies [@cormier2013; @gur2018; @becker2009]. Nevertheless, it is paramount that we investigate and describe the effects of language deprivation among deaf adults on narrative competence more in depth. This will help us understand whether the loss of systematic and frequent language input in the early years of life affect how deaf signers produce a narrative in their adulthood. Given this, the present study will focus on deaf adult signers with language deprivation and examine their reference tracking strategies in signed narratives.

Sign Languages are natural languages of the Deaf communities all around the world. They are independent languages, which do not derive from majority spoken language(s) but emerge naturally when deaf people form a community. Turkish Sign Language (Türk İşaret Dili -- TİD) is and has been the officially recognized language of the Deaf community in Turkey since 2005 [see @nuhbalaoglu2018 for a brief review of the historical development of the language]. Deaf or hearing children acquiring a SL within a family where there is one or more deaf caregivers display the same developmental milestones as hearing children acquiring a spoken language in a hearing family [@chamberlain_language_1999; @meier1990]. However, reports indicate that over 90% of all deaf individuals are born into hearing families who do not know how to sign (i.e., those with language deprivation) [@woll2013]. This means that the auditory modality used in the household is not accessible to many deaf children. That is why the age of acquisition of a first functional language varies greatly among deaf children, with many of whom being deprived of their first languages typically until after age 7 when they start formal education in a deaf school. Despite this, many of the early intervention programs for late signing children prioritize spoken language development and discourage the use of a sign language [@kemaloglu2012]. @ministryofnationaleducation2020 identifies 72 deaf schools in Turkey for the K-12 level. Even when deaf children with hearing parents start their formal education in one of these schools, they do not receive proper instruction in TİD since the teachers themselves are mainly hearing and only have a very rudimentary knowledge of a sign language [@ilkbasaran2015]. First language deprivation or delay has been found to have negative effects on certain cognitive and linguistic measures among late signing children. To name a few, delayed exposure to a first functional language is reported to negatively affect deaf children and adults' morphosyntactic production and processing [@kayabasi2020; @boudreault2006; @newport1990; @cheng2021; @karadoller2017; @cormier2012] and narrative abilities [e.g., @becker2009]. In our overall research, we are concerned with the effects of this late exposure in the ultimate adult forms of TİD users.

@@

Natural languages can be systematically used to keep track of previous referents in the discourse and mark the changing cognitive statuses of the REs (i.e., from being active to less active in the addressee's mind) to create a more coherent narrative. Among the theoretical frameworks that aim to explain that different linguistic forms are selected in accordance with the addressee's mental representations of the referents in the discourse is the Givenness Hierarchy [@gundel1993]. According to this model, given or familiar information is expressed in more attenuated forms such as pronominal use and null marking since the mental status of the expressed referent is already at the center of attention, as indicated by the present contextual information [@chafe1976]. The Givenness Hierarchy is akin to other previously proposed models of reference control within the topicality and saliency paradigm such as the Accessibility Hierarchy [@ariel1990], Topic Continuity [@givon1983], and Familiarity Scale [@prince1981]. @gundel1993 argued that the distribution of forms is not random but abides by the Gricean Maxim of Quantity [@grice1975]. That is to say, individuals must make the most informative and succinct choice possible to retain economy of form. When multiple expressions are available for the same referent as in an utterance like (\ref{ex:1}), phonetically simplest form is selected with the least redundant information based on an economy principle. In the utterance, REs for "Eddie" are given in bold. All three REs (nominal, pronominal, and null marking) refer to the same entity; however, the quantity of marking decreases as the referent's accessibility increases. On this ground, for highly accessible referents (i.e., those that are highly active in the mind), linguistic forms with the least phonetic content are selected for the least amount of processing effort.

```{=tex}
\begin{exe}
    \ex \label{ex:1}
    So \textbf{Eddie} turned around. \hfill Williams (1988; p. 343, ex. 7) \\
\textbf{He} said, "youse got a problem?" \\
"Yeah, we want you," they say.
So-$\pmb{\emptyset}$ walked right up to them \\
\end{exe}
```
It is also possible to categorize REs into three according to their discourse status [@gullberg2006]. A referent can be introduced or mentioned for the first time, maintained, that is, continued across two or more clauses, or re-introduced into the discourse following a topic shift. As a result, interlocutors mark some REs differently than others based on the principles of reference accessibility. For spoken languages, this observation is well established [@ariel1990; @givon1983; @gundel1993; @debreslioska2013; @ahrenholz2005; @chafe1976; @hickmann1996; @azar2015]. These research indicate that the introduction and re-introduction of a referent are often accomplished by full noun phrases which can be commonly modified by an adjective or demonstrative. Maintenance, on the other hand, is typically marked with pronominals and null markers since the continued referent is still highly accessible to the addressee.

However, this particular observation is not peculiar to spoken languages. There is an emerging body of literature that reports the distribution of REs in signed narratives [@czubek2017; @swabey2002; @frederiksen2016; @perniss2015; @morgan2005; @hodge2019; @ahlgren1994]. Building on @gundel1993's work, @swabey2002 demonstrated that, like other spoken languages such as Russian, Chinese, and Japanese, American Sign Language (ASL) signers use bare nouns for introduced and re-introduced discourse contexts whereas null marking is used for referents which are "in-focus" or very accessible to the addressee. Overt REs like pronouns or nouns are disprefered for such referents. @swabey2002 concluded that "native" signers in her study abided by the Gricean maxim of quantity and were thus economic in their reference tracking. Similar findings have been reported for German Sign Language (Deutsche Gebärdensprache - DGS) [@perniss2015] and British Sign Language (BSL) [@morgan2005]. It is important to note, however, that these research either collapsed native and late signers into a single group and labeled them as "native-like" or explicitly compared native deaf signers with late deaf signers (with language deprivation) in terms of narrative competence.

@@

The linguistic inventory for discourse referents can include nominal constructions (bare nouns, fingerspelled nouns, and modified nouns), null markers (plain verb, agreement verb, and constructed action), classifier predicates (whole entity classifiers, body part classifiers, handling classifiers, and extension classifiers), and overt pronouns (pronominal index or IX) [@frederiksen2016]. This interim section describes such constructions attested for sign languages under four main headings (nominals, pronouns, classifiers, and null marking).

@@

In TİD and many other sign languages, it is possible to use bare noun phrases as in (\ref{ex:6a}) as well as fingerspelled nouns as in (\ref{ex:6c}). Fingerspelling refers to the use of the manual alphabet which is derived or borrowed from a surrounding written language [@kubus2008]. Nouns can also be modified by classifiers as in (\ref{ex:6b}) [@frederiksen2016] or by a prenominal or postnominal IX functioning as a determiner or a demonstrative as in (\ref{ex:6d}) [@maclaughlin1997].

### Pronouns

The IX sign (extended index finger) has various functions in sign languages, one of which is the pronominal use which refers to established nominals by pointing to a previously assigned abstract or physical locus in the signing space when re-introducing a topic in discourse [@emmorey1996]. The same IX function has also been identified for TİD [@sevinc2006]. The example in (\ref{ex:7}) illustrates a spatially-anchored pronominal reference for topic re-introduction. These signs are thus rarely used for the first mention of an entity [@perniss2015]. Other interpretations consider pointing signs without noun phrases as examples of demonstratives and not personal pronouns [@koulidobrova2016]. It is also important to mention here that in some spoken languages (e.g., German and Dutch), demonstratives can also refer to persons [@bosch2007] and behave similarly to personal pronouns. However, for the purposes of the present study, we will refer to IX constructions as pronominal IX and particularly examine their non-deictic or personal pronoun realizations as a possible discourse strategy to mainly mark topic shift among native and late signers of TİD.

@@

Classifier predicates are highly complex morphological constructions which are very common in sign languages [@supalla1986]. These are manually signed iconic representations of animate or inanimate objects, and they embody several properties pertaining to the nominal referent including its location in the signing space, movement, path, size, and shape [@talmy2003]. As in most sign languages hitherto studied, classifier constructions are also commonly found in TİD [see @dikyuva2015 for a review]. Following the classification of @engberg-pedersen1993, we distinguish between four types of classifier constructions. Whole entity classifiers (WECL) represent the entirety of any referent whereas body part classifiers (BPCL) only reflect a specific part of an animate entity (e.g., limbs, head, tongue, teeth, and so forth). Handling classifiers (HCL[^1]) occur with transitive verbs and depict how an object is handled or manipulated by an agent. Extension classifiers (ExtCL), on the other hand, denote the size and shape of an entity by following a trajectory to trace its shape. We exemplify these constructions in (\ref{ex:8}).

[^1]: As discussed before, an HCL is transitive in nature and depicts the handling action performed by ananimate agent. This is why most research tend to consider HCLs as an extension to constructed action [@dudis2004; @cormier2013].

```{=tex}
\begin{multicols}{2}
\begin{exe}
\ex
\begin{xlist}
\ex \label{ex:6a} Full noun \\\glll
\includegraphics[width=50pt]{pictures/1a_1.png}\\ \textsc{mouse\footnotemark} \\ \\ 
\ex \label{ex:6b} Noun modified by classifier \\\glll
\includegraphics[width=50pt]{pictures/1c_2.png} \includegraphics[width=50pt]{pictures/1c_1.png} \\ \textsc{soap} \textsc{soap.cl} \\ \\ 
\ex \label{ex:6c} Fingerspelled noun\\\glll 
\includegraphics[width=50pt]{pictures/1b_1.png} \includegraphics[width=50pt]{pictures/1b_2.png} \includegraphics[width=50pt]{pictures/1b_3.png}\\ \textsc{t-} \textsc{o-} \textsc{m} \\ \\
\ex \label{ex:6d} Noun modified by index (\textsc{IX})\\\glll 
\includegraphics[width=50pt]{pictures/1d_1.png} \includegraphics[width=50pt]{pictures/1d_2.png}\\ \textsc{ix} \textsc{cat} \\ \\
\end{xlist}
\end{exe}
\end{multicols}
\ea 
\ea \label{ex:7} Pronominal \\\glll
\includegraphics[width=50pt]{pictures/2_1.png} \includegraphics[width=50pt]{pictures/2_2.png} \\ \textsc{ix} \textsc{sick} \\ \\
\z \z
\clearpage
\begin{multicols}{4}
\begin{exe}
\ex \label{ex:8}
\begin{xlist}
\ex Whole entity \\\glll 
\includegraphics[width=50pt]{pictures/3e.png} \\ \textsc{fish.fall}\\ \\
\ex Body part \\\glll
\includegraphics[width=50pt]{pictures/3b.png} \\ \textsc{walk} \\ \\ 
\ex Handling \\\glll 
\includegraphics[width=50pt]{pictures/3c.png} \\ \textsc{hold} \\ \\
\ex Extension \\\glll 
\includegraphics[width=50pt]{pictures/3d_1.png} \includegraphics[width=50pt]{pictures/3a.png}\\ \textsc{plank} {\textsc{dh: mouse}} \\ {} {\textsc{ndh: tunnel}}\\ \\
\end{xlist}
\end{exe}
\end{multicols}
\begin{multicols}{3}
\begin{exe}
\ex
\begin{xlist}
\ex \label{ex:9a} Constructed action \\\glll 
\includegraphics[width=50pt]{pictures/4c.png} \\ \textsc{struggle} \\ \\

\ex \label{ex:9b} Plain verb  \\\glll
\includegraphics[width=50pt]{pictures/4a_2.png}  \\ \textsc{enter}\\ \\

\ex \label{ex:9c} Agreement verb  \\\glll 
\includegraphics[width=50pt]{pictures/4b_1.png} 
\includegraphics[width=50pt]{pictures/4b_2.png} \\ {\textsc{$_{1}$hit$_{3}$}} {} \\ \\ 
\end{xlist}
\end{exe}
\end{multicols}
\footnotetext{We use small capitals to indicate sign glosses.}
```
### Null marking

>>>>

---
editor_options:
  chunk_output_type: console
title: Reference Tracking Strategies of Deaf Adult Signers in Turkish Sign Language
abstract: |
  We investigated the reference tracking strategies among deaf adults with signing deaf parents (DoD) and adult deaf signers with non-signing hearing parents or caregivers (DoH) who had late exposure to Turkish Sign Language (TİD). Consistent with the theories of saliency and referential accessibility, regardless of their acquisiton groups or parental hearing status, signers mainly used nominals and extension classifiers for introducing referents. To maintain a referent from the previous clause, the signers used zero anaphora (e.g., constructed action, agreement and plain verbs). For topic shifts or re-introduced contexts, nominals and pronouns were chiefly favored although we observed very little pronominal use. As for the effect of native acquisition from deaf caregivers, we only report limited over-redundancy for DoH signers who used zero anaphora less compared to DoD especially for introduced and maintained contexts. We conclude that DoH signers are still able to achieve native-like competency in terms of reference tracking in simple narratives but DoD signers utilize the spatial affordances of the visual-spatial modality better than DoH signers only to a certain extent.
keywords: "Referential accessibility, signed narratives, pragmatic competence, delayed language acquisition, Turkish Sign Language"
journal: "Journal of Pragmatics"
date: "`r Sys.Date()`"
bibliography: ./bibliography/references.bib
times: true
numbersections: trueand w
classoption: "review"
csl: ./bibliography/elsevier-harvard.csl
output: rticles::elsevier_article
header-includes:
   - \usepackage{floatrow}
   - \usepackage{hyperref}
   - \usepackage{enumitem}  
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \floatplacement{table}{H}
   - \floatsetup[figure]{capposition=top}
   - \floatsetup[table]{capposition=top}
   - \usepackage{times}
   - \usepackage{smartdiagram}
   - \usepackage{ragged2e}
   - \usepackage{tikz}
   - \usepackage{xcolor}
   - \usepackage[utf8]{inputenc}
   - \usepackage[T1]{fontenc}
   - \usepackage{graphicx}
   - \usepackage{./sty/langsci-gb4e}
   - \usepackage{caption}
   - \usepackage{multicol}
   - \setlength{\columnsep}{-2cm}
   - \usepackage{setspace}
---

@@

Narratives are extensive discourses that form an important part of human communication. In order to create a systematic and coherent organization of experiences, narrators must choose appropriate linguistic forms to refer to entities [@ariel1990; @givon1983; @gundel1993]. Certain forms such as nominal expressions are commonly used for the first mention of a referent in a discourse, and other forms such as pronouns and zero anaphora may track previously mentioned entities which could either be maintained across multiple clauses or re-introduced back into the discourse. The selection of these referring expressions (henceforth REs) is contingent on cognitive abilities, such that individuals must rely on the activation of the working memory to keep in mind previously referred expressions and long strings of information [@morgan2005; @bamberg1997]. Pragmatic abilities too play a pivotal role in the tracking of referents. The form with the least amount of redundant information must be selected based on the accessibility levels of the referent (i.e., how active a referent is in the addressee's mind given a point in discourse) [@ahn2019; @ariel1990]. Because of these cognitive and pragmatic necessities, it is not surprising to observe that producing a coherent narrative is one of the latest milestones in language acquisition among children [@hickmann1996]. As can be expected, there is a plethora of research examining how narrative competence is achieved in both monolingual and bilingual acquisition contexts among children and adults. Although narrative competence and referent selection have been studied extensively for typically developing speakers [@aksu-koc2015; @hickmann1996; @williams1988], deaf individuals who acquired a sign language in atypical learning contexts (i.e., deaf signers with delayed exposure to a sign language) have remained mostly understudied with an exception of a handful of studies [@cormier2013; @gur2018; @becker2009]. Nevertheless, it is paramount that we investigate and describe the effects of late exposure to a sign language among deaf adults growing up in a signing vs. non-signing environment on narrative competence more in depth. This will help us understand whether the lack of systematic and frequent language exposure in the early years of life affect how deaf signers produce a narrative in their adulthood. Given this, the present study will focus on two groups of signers and examine their reference tracking strategies in signed narratives: (i) deaf of deaf (DoD) adult signers (i.e., signers who have deaf caregivers and natively acquired a sign language from birth) and (ii) deaf of hearing (DoH) adult signers (i.e., signers who have hearing non-signing caregivers and state to have started acquiring a sign language later in life).

Sign languages are the natural languages of the Deaf communities all around the world. They are independent languages, which do not derive from majority spoken language(s) but emerge naturally when deaf people form a community. Turkish Sign Language (Türk İşaret Dili -- TİD) is and has been the officially recognized language of the Deaf community in Turkey since 2005 [see @nuhbalaoglu2018 for a brief review of the historical development of the language]. Deaf or hearing children acquiring a SL within a family where there is one or more deaf caregivers display the same developmental milestones as hearing children acquiring a spoken language in a hearing family [@chamberlain_language_1999; @meier1990]. However, reports indicate that over 90% of all deaf individuals are born into hearing families (DoH) who do not know how to sign [@woll2013]. This means that the auditory modality used in the household is not accessible to many deaf children. That is why the age of acquisition of a first functional language varies greatly among deaf children, with many of whom being deprived of their first languages typically until after age 7 when they start formal education in a deaf school, especially in the context of Turkey [@sari_analysis_2005]. Despite this, many of the early intervention programs for late signing children prioritize spoken language development and discourage the use of a sign language [@kemaloglu2012]. The @ministryofnationaleducation2020 identifies 72 deaf schools in Turkey for the K-12 level. Even when deaf children with hearing parents start their formal education in one of these schools, they do not receive proper instruction in TİD since the teachers themselves are mainly hearing and only have a very rudimentary knowledge of a sign language [@ilkbasaran2015; @sari_analysis_2005]. In some cases, it has been argued that late exposure to a sign language among certain DoH signers has negative effects on some cognitive and linguistic measures. To name a few, delayed exposure to a first functional language is reported to negatively affect deaf children and adults' morphosyntactic production and processing [@kayaba2022; @boudreault2006; @newport1990; @cheng2021; @karadoller2017; @cormier2012] and narrative abilities [e.g., @becker2009]. In our overall research, we are concerned with the effects of this late exposure in the ultimate adult forms of TİD users.

@@

Natural languages can be systematically used to keep track of previous referents in the discourse and mark the changing cognitive statuses of the REs (i.e., from being active to less active in the addressee's mind) to create a more coherent narrative. Among the theoretical frameworks that aim to explain that different linguistic forms are selected in accordance with the addressee's mental representations of the referents in the discourse is the Givenness Hierarchy [@gundel1993]. According to this model, given or familiar information is expressed in more attenuated forms such as pronouns and zero anaphora (i.e., there is no explicit mention of the referent) since the mental status of the expressed referent is already at the center of attention, as indicated by the present contextual information [@chafe1976]. The Givenness Hierarchy is akin to other previously proposed models of reference control within the topicality and saliency paradigm such as the Accessibility Hierarchy [@ariel1990], Topic Continuity [@givon1983], and Familiarity Scale [@prince1981]. @gundel1993 argued that the distribution of forms is not random but abides by the Gricean Maxim of Quantity [@grice1975]. That is to say, individuals must make the most informative and succinct choice possible to retain economy of form. When multiple expressions are available for the same referent as in an utterance like (\ref{ex:1}), phonetically simplest form is selected with the least redundant information based on an economy principle. In the utterance, REs for "Eddie" are given in bold. All three REs (nominal, pronominal, and zero anaphora) refer to the same entity; however, the quantity of marking decreases as the referent's accessibility increases. On this ground, for highly accessible referents (i.e., those that are highly active in the mind), linguistic forms with the least phonetic content are selected for the least amount of processing effort.

```{=tex}
\begin{exe}
    \ex \label{ex:1}
    So \textbf{Eddie} turned around. \hfill Williams (1988; p. 343, ex. 7) \\
\textbf{He} said, "youse got a problem?" \\
"Yeah, we want you," they say.
So-$\pmb{\emptyset}$ walked right up to them \\
\end{exe}
```
It is also possible to categorize REs into three according to their discourse status [@gullberg2006]. A referent can be introduced or mentioned for the first time, maintained, that is, continued across two or more clauses, or re-introduced into the discourse following a topic shift. As a result, interlocutors mark some REs differently than others based on the principles of reference accessibility. For spoken languages, this observation is well established [@ariel1990; @givon1983; @gundel1993; @debreslioska2013; @ahrenholz2005; @chafe1976; @hickmann1996; @azar2015]. These research indicate that the introduction and re-introduction of a referent are often accomplished by full noun phrases which can be commonly modified by an adjective or demonstrative. Maintenance, on the other hand, is typically marked with pronominals and zero anaphora since the continued referent is still highly accessible to the addressee.

However, this particular observation is not peculiar to spoken languages. There is an emerging body of literature that reports the distribution of REs in signed narratives [@czubek2017; @swabey2002; @frederiksen2016; @perniss2015; @morgan2005; @hodge2019; @ahlgren1994]. Building on the work by @gundel1993, @swabey2002 demonstrated that, like other languages such as Russian, Chinese, and Japanese, American Sign Language (ASL) signers use bare nouns for introduced and re-introduced discourse contexts whereas zero anaphora is used for referents which are "in-focus" or very accessible to the addressee. Overt REs like pronouns or nouns are disprefered for such referents. @swabey2002 concluded that "native" signers in her study abided by the Gricean maxim of quantity and thus used more economic forms of reference tracking. Similar findings have been reported for German Sign Language (Deutsche Gebärdensprache - DGS) [@perniss2015] and British Sign Language (BSL) [@morgan2005]. It is important to note, however, that this research either collapsed DoD and DoH signers into a single group and labeled them as "native-like" or explicitly compared native deaf signers with late deaf signers (with delayed exposure to a sign language) in terms of narrative competence.

@@

The linguistic inventory for discourse referents can include nominal constructions (bare nouns, fingerspelled nouns, and modified nouns), null or zero anaphors\footnote{Zero anaphora does not refer to the lack of a referent. It only alludes to the non-presence of a more explicit anaphoric form for the referent such as a nominal, pronominal, or classifiers.} (predicated by a plain verb, an agreement verb, or constructed action), classifier predicates (whole entity classifiers, body part classifiers, handling classifiers, and extension classifiers), and overt pronouns (pronominal index or IX) [@frederiksen2016]. This interim section describes such forms attested for sign languages under four main headings (nominals, pronouns, classifiers, and zero anaphora).

@@

In TİD and many other sign languages, it is possible to use bare nouns as in \ref{ex:2a} as well as fingerspelled nouns as in \ref{ex:2b}. Fingerspelling refers to the use of the manual aphabet to spell all or part of the letters of a word borrowed from a surrounding written language [@kubus2008]. Nouns can also be modified by a prenominal or postnominal extended index finger pointing sign, glossed as IX, functioning as a determiner or a demonstrative as in \ref{ex:2c} [@maclaughlin1997] or by a modifying classifier within the same NP as in \ref{ex:2d} [@frederiksen2016].\footnote{Spoken languages use pronouns, determines and demonstratives where sign languages use a pointing sign, usually glossed as IX, which assumes similar functions (Lillo-Martin and Klima, 1990). In addition, the pointing sign is used to referentially anchor a referent to a location in signing space.}

```{=tex}
\newpage
\begin{exe}
\ex
\begin{xlist}
\ex \label{ex:2a} Full noun \\\glll
{} \includegraphics[width=35pt]{pictures/p1.png} {} \\ \textbf{Gloss:} \textsc{mouse\footnotemark} \textsc{sick} \\ 
\textbf{Status:} Intro. \\
\glt \textbf{Context:} Narration 6, Event 1. Here, the signer introduces the mouse, Jerry, at the beginning of the narration with a bare noun.
\glt \textbf{Translation:} There is a sick mouse.
\vspace{0.35cm}

\ex \label{ex:2b} Fingerspelled noun \\\glll
{} \includegraphics[width=40pt]{pictures/p2.1.png} \includegraphics[width=40pt]{pictures/p2.2.png} \includegraphics[width=40pt]{pictures/p2.3.png} {} \\ \textbf{Gloss:} \textsc{t-} \textsc{-o-} \textsc{-m} \textsc{come.back} \\ 
\textbf{Status:} Re-intro. \\
\glt \textbf{Context:} Narration 7, Event 2. The signer introduces Tom two sentences before. She introduces the black cat in the preceding clause and then here re-introduces Tom by using fingerspelling.
\glt \textbf{Translation:} Tom comes back.
\vspace{0.35cm}

\ex \label{ex:2c} Noun modified by index (IX) \\\glll
{} \includegraphics[width=45pt]{pictures/p3.1.png} \includegraphics[width=40pt]{pictures/p3.2.png} {} {} {} \\ \textbf{Gloss:} \textsc{ix} \textsc{cat} \textsc{pipe.extend} \textsc{go.through} \textsc{exit} \\ 
\textbf{Status:} Re-intro. \\
\glt \textbf{Context:} Narration 1, Event 6. The signer introduces the cat in the first clause of the first event and re-introduces it several times including this last event of the narrative.
\glt \textbf{Translation:} The cat goes through and exits the thin pipe.
\vspace{0.35cm}

\ex \label{ex:2d} Noun modified by classifier (CL) \\\glll
{} \includegraphics[width=35pt]{pictures/p4.1.png} \includegraphics[width=35pt]{pictures/p4.2.png} \\ \textbf{Gloss:} \textsc{soap} \textsc{soap.cl} \\ 
\textbf{Status:} Intro. \\
\glt \textbf{Context:} Narration 5, Event 3. Here the signer introduces the soap to the discourse. The signer produces the lexical sign, \textsc{soap}, plus a classifier, \textsc{soap.cl}.
\glt \textbf{Translation:} The soap, which is like this.
\end{xlist}
\end{exe}

\footnotetext{We use small capitals to indicate sign glosses.}
```
### Referential Use of Space and Pronouns

In addition to being a modifier of a noun, the IX sign can also be pronominal by referring to established nominals through pointing to a previously assigned abstract or physical locus in the signing space when re-introducing a topic in discourse [@emmorey1996]. This function has also been identified for TİD [@sevinc2006]. The example in \ref{ex:3} illustrates this spatially-anchored pronominal use.

```{=tex}
\ea 
\ea \label{ex:3} Pronominal \\\glll
{} \includegraphics[width=35pt]{pictures/p5.png} {} {} {} {} \\ \textbf{Gloss:} \textsc{ix} \textsc{look} \textsc{cat.sad(constructed action)} \textsc{sad} \textsc{sick} \\ 
\textbf{Status:} Re-intro. \\
\glt \textbf{Context:} Narration 9, Event 1: The signer introduces the mouse then signs an action of the cat. Here, he re-introduces the mouse with a pointing sign that functions as a pronoun.
\glt \textbf{Translation:} It (the mouse) looks sad and sick.
\z \z
```
These pronominal signs are rarely used for the first mention of an entity [@perniss2015]. Other researchers consider pointing signs without noun phrases as examples of demonstratives and not personal pronouns [@koulidobrova2016]. It is also important to mention here that in some spoken languages (e.g., German and Dutch), demonstratives can also refer to persons [@bosch2007] and behave similarly to personal pronouns. However, for the purposes of the present study, we will refer to IX signs as pronominal IX and particularly examine their non-deictic or personal pronoun realizations as a possible discourse strategy to mainly mark topic shift (re-introduction) among DoD and DoH signers of TİD.

@@

In sign languages, a verbal classifier morpheme is morpho-phonologically encoded by a handshape. This handshape is articulated co-temporally with a movement that encodes the verbal meaning [@supalla1990; @benedicto2004]. Semantically, the classifiers denote sets of objects based on criteria such as animacy, size, shape, body-part, texture and sometimes function [@talmy2003]. The classifier articulated simultaneously with a movement is called a classifier predicate and the sentence in which such a predicate is used is usually called a classifier construction in sign language research. As in most sign languages hitherto studied, classifier constructions are also commonly found in TİD [see @dikyuva2015 for a review]. Following the classification of [@engberg-pedersen1993], in TİD, we distinguish between four types of classifier handshapes. Whole entity classifiers (WECL) represent the entirety of a referent (e.g. an entire flat horizontal entity, or an upright entity) (e.g., \ref{ex:4a}) whereas body part classifiers (BPCL) represent a specific part of an animate entity (e.g., limbs, head, mouth, tongue, or teeth, etc.) (e.g.,\ref{ex:4b}). Handling classifiers (HCL) occur with transitive verbal roots and represent both the size, shape and surface of an object and how that object is handled or manipulated by an agent (e.g., \ref{ex:4c}). Extension classifiers (ExtCL), on the other hand, represent the size and shape of an entity by following a trajectory to trace its shape (e.g., \ref{ex:4d}). We exemplify these handshapes with the discourse status of the referent they track and a description of the previous discourse when available.

```{=tex}
\begin{exe}
\ex
\begin{xlist}
\ex \label{ex:4a} Whole entity classifier  (WECL) \\\glll
{} \includegraphics[width=40pt]{pictures/p6.png} \\ \textbf{Gloss:} \textsc{two.mice.go.together} \\ 
\textbf{Status:} Maintain. \\
\glt \textbf{Context:} Narration 4, Event 2: The signer introduces Jerry and the robot mouse in the preceding clause. She maintains reference to them by using a whole entity classifier on each hand. 
\glt \textbf{Translation:} (Jerry and the robot mouse from the previous context) go together.
\vspace{0.35cm}

\ex \label{ex:4b} Body part classifier (BPCL) \\\glll
{} \includegraphics[width=40pt]{pictures/p7.png} \\ \textbf{Gloss:} \textsc{cat.move.thin.legs} \\ 
\textbf{Status:} Maintain. \\
\glt \textbf{Context:} Narration 1, Event 6: The signer re-introduces the cat three clauses before and maintains reference to him in the two clauses before and here with a body part classifier that refers to the legs of the cat.
\glt \textbf{Translation:} (The cat) walks (i.e. moving his thin legs forward).
\vspace{0.35cm}

\ex \label{ex:4c} Handling classifier (HCL) \\\glll
{} {} {} {} {} \includegraphics[width=40pt]{pictures/p8.png} \\ \textbf{Gloss:} {} \textsc{rope see-3a} \textsc{soon} \textsc{go} \textsc{cat.pull.rope} \\ 
\textbf{Status:} Maintain. \\
\glt \textbf{Context:} Narration 5, Event 2: The signer introduces Tom three clauses before. She maintains reference to him in the preceding two clauses and here she still maintains reference to him by using a handling classifier on both hands.
\glt \textbf{Translation:} When (he/Tom) sees the rope, (he) soon goes to pull it.
\vspace{0.35cm}

\ex \label{ex:4d} Extension classifier (ExtCL) \\\glll
{} \includegraphics[width=80pt]{pictures/p9.png} \\ \textbf{Gloss:} \textsc{cat.extend} \\ 
\textbf{Status:} Maintain. \\
\glt \textbf{Context:} Narration 1, Event 6: The signer re-introduces Tom two clauses before and she maintains reference to him with constructed action in the previous clause and here with an extension classifier.
\glt \textbf{Translation:} (The cat) extends thin.
\end{xlist}
\end{exe}
```
### Zero Anaphora

Parallel to spoken languages, many sign languages have been reported to allow null subjects and/or objects and make use of null arguments when referring to highly accessible referents that are still active in the mental representation of the addressee. In signed discourse, this can be achieved by employing certain verbal constructions or constructed action (CA). The latter construction exemplified in \ref{ex:5a} refers to cases in which the signers engage with the space in front of themselves as if they are characters in the story interacting with objects and other characters. This is often achieved with the signer taking on the role of a referent in the narrative and performing the actions [@metzger1995]. CA is often marked by certain non-manual cues such as the facial expression and the breaking of the eye contact. This is to say, the narrator's head, face, torso, arms, and as an optional marker, hands represent those of the associated referent(s) [@smith2014]. As for the verbal constructions, to identify the predicate types in ASL, @padden1986 suggested a triadic categorization of verbs in sign languages: plain verbs, agreement verbs, and spatial verbs. According to this classification, plain verbs (as in \ref{ex:5b}) do not agree with the subject or the object. This means that they are not inflected for number or person. This contrasts with agreeing verbs (as in \ref{ex:5c}) the starting or the ending point of which can be modified to specific loci for person and number agreement while spatial verbs (as in \textsc{ix book put} 'I placed the book there') agree with locative arguments but like plain verbs, they are not inflected for number or person. TİD too licenses null arguments. Following this classification, @sevinc2006 argued that although it is common in TİD to drop arguments, there were certain limitations on the null marking of the arguments for agreement and plain verbs.
